Creating More Meaningful Explanations for Audio Deepfake Detection

### 2025-03-05 draft.tex
more detail in method section about the model architecture + new figures

% To address this problem and due to a lack of metrics for assessing the aggregated average value of local explanations, we introduce two metrics: \textit{trust} and \textit{importance}. These metrics allow for an aggregate evaluation of a large number of generated LIME explanations on a per-feature basis, enabling us to assess the value of the selected feature set for producing sensible explanations.

We conducted a wide range of experiments to evaluate the performance of our hybrid audio deepfake detection approach on the \textit{In-The-Wild} dataset. The experimental results demonstrate that our model performs effectively compared to state-of-the-art methods. The evaluation of the generated explanations using our proposed metrics also identifies which features have the highest impact on the model's overall predictions. The contributions of this paper are twofold:
\begin{itemize}
\item We propose an interpretable hybrid deep learning model to identify synthetic voices using perceptible features.
% \item Our method can explain predictions, and based on our introduced evaluation metrics, the explanations can highlight the impact of specific features.
\item We propose two metrics in order to perform aggregate evaluations of a large explanations from our LIME-based method, in order to assess how useful the average explanation is.
\end{itemize}

@@

%Since the use of generative methods for the creation of synthetic voices becomes more widespread, the need for reliable and usable detection methods for the protection of the security of people and businesses grows. In particular, the rise of deepfake technology has led to concerns about its potential misuse. There are several areas where the misuse occurs including politics, entertainment and national security by, for instance, potentially allowing malicious actors to create fake audio recordings that appear to be genuine statements made by public figures or to create genuine-seeming recordings of events that never happened, which could have significant consequences if used to spread misinformation, from defaming individuals to creating political tension\cite{veerasamy2022rising,albahar_deepfakes_2005}.

%we proposed a hybrid explainable fake voice detection model leveraging perceptible features extracted from the voice signal. Since the dimension of the different features are not same (i.e., multidimensional), we first pass through the every single feature values through individual sub-models. In other words, the hybrid model has first individual sub-model of every features and the concatenation of the similar representation is going through a terminus model. At the output layer of such sub-model, we have similar representation of the features values with same dimension. Next the terminus model process the concatenation of the output from all sub-models.

%we make a distinction of two main categories of input features for a synthetic voice detection model: perceptible and imperceptible features. Perceptible features are features that can be perceived by the human ear, often vocal qualities such as jitter, shimmer or pitch fluctuation that have a wide range of uses even outside of audio classification such as diagnosis of disease \cite{chaiwongyen_deepfake-speech_2023}, while imperceptible features are typically out of the range of human hearing, and may not directly reflect a vocal quality, otherwise referred to as "speaker-independent" features~\cite{liu_hidden--wave_2023}. We use a new combination of features for our method, with an emphasis being placed on perceptible features, as we believe these to be the most likely to be understood by the layperson.

%The contributions of this work are in summary: Our hybrid model-architecture and modified LIME-based method for generating explanations of classifications of audio voice samples as synthetic or bona-fide and the two metrics that we introduce in order to assess the impact of individual features over numerous local explanations.
	%\begin{itemize}
	%	\item §\ref{sec:related} Related Work, where we discuss the most relevant related research in the field, in order to create a picture of the current state-of-the-art
	%	\item §\ref{sec:method} Method, where we describe the general model implementation, as well as the implementation of our LIME-based method for generating explanations in more detail.
	%	\item §\ref{sec:experimental_results} Experimental Results, where we discuss how we performed our experiments, how our hybrid model performs, as well as use our proposed metrics in order to assess the feature impact in the average explanation.
        %\item §\ref{sec:discussion} Discussion, where we discuss the implication of the results of our experiments, including the potential of the explanations as well as the	shortcomings and limitations of our method.
	%\end{itemize}

@@

%\section{Related Work}	\label{sec:related}
%There has been a significant attention to detect fake voices in recent years. This section presents related research which makes use of hand-crafted features, both perceptible and imperceptible features. In the end, we present the effort of the research community to make fake voice identification methods explainable.

%\subsection{Synthetic Voice Detection with Imperceptible Features}
%The most common approach in this field makes use of either learned or hand-crafted imperceptible features. Examples include spectrographic features such as the mel-spectrogram and its hand-crafted derivative the mel-frequency cepstral coefficients (MFCCs), both of which are present in our method due to their widespread successful use.

%The work of Anagha et al. \cite{anagha_audio_2023} is an example that makes use of mel-spectrograms in combination with an architecture based on convolutional neural networks, achieving well-performing results on the ASVSpoof2019 dataset \cite{wang_asvspoof_2020}. A further work from Yan et al. \cite{yan_initial_2022} makes use of MFCCs, in combination with other hand-crafted imperceptible features such as linear frequency cepstral coefficients (LFCCs) to not only classify audio samples as synthetic or authentic, but also to correctly classify the vocoder used to produce the sample with nearly perfect accuracy on a handcrafted dataset.

%Other researchers, such as in the work of Qais et al. \cite{qais_deepfake_2022} make use of Fourier transforms, such as the short term Fourier transform (STFT). The aforementioned example uses these features combined with others such as the aforementioned mel-spectrogram based features. They also pair these features with a convolutional architecture and achieve notably accurate results on the ASVSpoof2017 dataset \cite{delgado_asvspoof_2018}. With their work, Müller et al. \cite{muller_does_2022} introduce the In-the-Wild dataset, in an attempt to answer the question if -- and to what extent -- previously proposed methods of deepfake voice detection can generalize. They test a range of features and architectures on their dataset, concluding that there is still work to be done in the field, due to a generally lackluster performance.

%Yang et al. \cite{yang_robust_2024} demonstrate a comparison of multiple features, as well as a feature selection method, with the goal of maximizing model efficiency. They run experiments with several hand-crafted features as well as learned features. They train and test using three datasets, ASVSpoof2019 \cite{wang_asvspoof_2020}, ASV2021 \cite{liu_asvspoof_2023} and In-the-Wild \cite{muller_does_2022}, presenting their results for each feature, as well as the results of their selection and classification fusion methods.

%The work of Ranjan et al. \cite{ranjan_statnet_2022} introduces a deep convolutional network that is designed not only for spoof detection, but also for the purpose of source detection. They perform experiments on several datasets, including ASVSpoof2019 \cite{wang_asvspoof_2020}, the FOR-Norm dataset \cite{reimao_for_2019} and the In-the-Wild dataset \cite{muller_does_2022}. In the experiments run on each dataset, their method was able to achieve high evaluation performance, but they were not able to achieve high-performance in cross-dataset evaluations, indicating a lack of generalizability.

%Yi et al. \cite{yi_audio_2023} present in their work a comprehensive comparison of features and model architectures used for detection methods, giving results from both the ASVSpoof2019 \cite{wang_asvspoof_2020} and In-the-Wild \cite{muller_does_2022} datasets, once again demonstrating a lack of generalizability of existing methods, consistent across various kinds of features and architectures.

%\subsection{Synthetic Voice Detection With Perceptible Features}
%There is also a smaller but still notable body of work encompassing the use of perceptible features for the detection of synthetic voices, making use of a variety of features, some of which were selected for use in our experiments.

%The work of Barrington et al.~\cite{barrington_single_2023} explores the idea of classification using perceptible features, highlighting their potential for improving explainability in the space of deepfake audio detection, although they only implement a classifier and not an explainer. They do, however, also demonstrate a drop in performance when using perceptible features in comparison with imperceptible hand-crafted and deep-learning features in their experiments, which were performed on a combination of synthetic and real audio datasets.

%Chaiwongyen et al.~\cite{chaiwongyen_contribution_2022,chaiwongyen_deepfake-speech_2023} also approach using perceptible features for classification in their works. They developed a perceptron architecture with a single hidden layer and trained and tested using the dataset of the ADD2022 Challenge~\cite{yi_add_2024}, resulting in lackluster performance in 2022, with better results using an expanded and improved feature set in 2023. Li et al.~\cite{li_comparative_2022} also approach perceptible features in their work, combining them with imperceptible (referred to as physical in the work) features and using various neural networks in their experiments. With their combination of perceptible and imperceptible features, they were able to produce the best performance results of their experiments in comparison to experiments run with only perceptible or imperceptible features, which were performed as well on the dataset of the ASV2022 Challenge~\cite{yi_add_2024}.

%\subsection{Explainable Models}
%There have also been some efforts in the previous research to pair synthetic voice detection models with explainable methods.

%One previous example of an implementation of an explainable model for use in audio deepfake detection came from Ge et al.~\cite{ge_explaining_2024}, who explain feature influence on models using the SHAP (SHapley Additive exPlanations) method. They apply this method using log-scaled power spectrograms as a feature, training and testing using the datasets from the ASV2019 challenge~\cite{wang_asvspoof_2020}, they are able to determine and graphically represent areas of importance on the spectrogram, as well as globally summarize the SHAP-values.

%Another example is presented by Haq et al.~\cite{haq_multimodal_2023}, who use the changes in emotional state as an input feature and represent "unlikely" changes on a graph so that it can be understood by an eventual end user. In this case, they achieved their results by combining the output of fake video and fake audio classifiers in order to produce a final classification for a video sample with audio. Testing against the presidential deepfake dataset \cite{sankaranarayanan_presidential_nodate}, they achieve impressive results in comparison to the standing benchmark on the dataset at the time.
 \begin{figure}[t]
    \centering
    \includegraphics[width=0.75\linewidth]{images/Picture1.png}
    \caption{A high level building-block of our proposed hybrid fake voice identification model, where \(m\) and \(k\) represent the dimensionality of the given input feature.}
    \label{fig:overview}
\end{figure}
\section{Method}\label{sec:method}
Our proposed fake voice identification method consists of three major components including feature extractor, hybrid detection model, and the generation of explanations by modified LIME. After extracting features, we pass them individually through sub-models to overcome the multidimensionality problem. Then the concatenation of the outputs from the output layers of the sub-models are then passed through the \textit{terminus} models. For generating explanations, we make use of the feature values returned from the output layers of the sub-models, which has same representation of every feature.

@@

%Several features were used for the final version of the black-box model. Some are features widely used in research involving the detection of synthetic while others are more specific and are cited accordingly. The features that are not summarized for a whole sample are extracted using a sliding window method, preventing a loss of fidelity that can be caused by compression of the feature to a standard size. The features we considered perceptible are as follows:

@@

\subsection{Hybrid Fake Voice Detection Model}
%We created, trained and evaluated the model used for our experiments using the tools available in the TensorFlow and Keras libraries \cite{tensorflow2015-whitepaper}. This section will go into further detail about the experimental setup for the black-box model as well as the way in which explanations were created and summarized.

Fig.~\ref{fig:overview} depicts the overall architecture of our hybrid deep learning model for fake voice identification. We propose the architecture of our hybrid fake voice detection model in a way that allows features with different dimensionality, regardless of whether the individual features have the same shape. To achieve this, we hypothesize to have individual separate model for each feature and we reoffered this as \textit{sub-model}. The main objective for having the individual sub-model is to have similar representation of the features so that we can use them for generating understandable explanations. The structure of sub-model can be different based the dimension of the extracted features. For example, the dimension of convolution layers and max pooling layer would be different based on the dimension of the features. However, followed by the concatenations of the output of all sub-models, we proposed a terminus models. To have the better performance, we tried different alternative structures of the terminus models including single layer perceptron, convolutional neural networks and multi-layer perceptron. The figure (Fig.~\ref{fig:overview}) has CNN based terminus. However, it can be a single layer perceptron or multilayer perceptron also. We discuss in the experiments section about our experimental settings.

The sub-models begin with an input layer, taking either a vector or matrix depending on the dimensionality of the feature to which the model is tailored. The inputs are then passed into either a series of convolutional layers if the input feature is in the form of a vector or alternating convolutional and pooling layers if the input feature is in the form of a matrix. We make use of convolution in our sub-models in order to increase the localized pattern detection capability within the sub-models. %Figure \ref{fig:term_model} is a generalized representation of the terminus model, which has a classic perceptron architecture, with a varying number of hidden layers, depending on the experiment.

%a separate model for each feature, with its own input and output layers (further referred to as sub-models), which are then concatenated together and processed though a model that pools into a final output value (which we will further refer to as the terminus model). The modular structure allows not only better performance and less memory use, due to the fact that fewer transformations are required for the input preparation, but also allows for more flexibility in the construction of the model, making it easy to add and remove features using pre-defined functions depending on feature type. This kind of architecture also provides an advantage to the generation of explanations, as only the terminus part of the hybrid model must be explained to interpret the importance of each input feature. Further clarification is in §\ref{sec:method_exp}.

%Figure \ref{fig:model_plot} depicts the meta-structure of our hybrid model, with each feature first being processed in its own separate sub-model before being concatenated into the terminus model at the end, which produces the final result. Figure \ref{fig:submodel} offers a deeper look into the general architecture of a singular sub-model in our method. The sub-models begin with an input layer, taking either a vector or matrix depending on the dimensionality of the feature to which the model is tailored. The inputs are then passed into alternating convolutional and pooling layers, in order to increase the localized pattern detection capability within the sub-models. Figure \ref{fig:term_model} is a generalized representation of the terminus model, which has a classic perceptron architecture, with a varying number of hidden layers, depending on the experiment.
\if false
    \begin{figure}[htbp]
        \begin{center}
            \includegraphics[width=0.7\textwidth]{images/model_fig.png}
            \caption{The general architecture for the models used in these experiments,
            containing \(n\) features. The features are processed in their tailored
            sub-models before being processed together in the terminus model.}
            \label{fig:model_plot}
        \end{center}
    \end{figure}
    \begin{figure}[htbp]
        \begin{center}
            \includegraphics[width=0.7\textwidth]{images/submodel.png}
            \caption{A generalized representation of a sub-model, with the input as either
            a vector or a matrix followed by alternating convolutional and pooling layers,
            with the output as a flattened vector. Here, \(n\) is the number of
            convolutional-pooling pairs.}
            \label{fig:submodel}
        \end{center}
    \end{figure}
    \begin{figure}[htbp]
        \begin{center}
            \includegraphics[width=0.7\textwidth]{images/term_model.png}
            \caption{A generalized representation of the terminus model paired with \(n\)
            sub-models and containing \(m\) hidden layers.}
            \label{fig:term_model}
        \end{center}
    \end{figure}
\fi

>>>>

We conducted a wide range of experiments to evaluate the performance of our hybrid audio deepfake detection approach on the \textit{In-The-Wild} dataset. The experimental results demonstrate that our model performs effectively compared to state-of-the-art methods. The evaluation of the generated explanations using our proposed metrics also identifies which features have the highest impact on the model's overall predictions. The contributions of this paper are twofold:
\begin{itemize}
\item We propose an interpretable hybrid deep learning model to identify synthetic voices using perceptible features.
\item We propose two metrics in order to perform aggregate evaluations of a large explanations from our LIME-based method, in order to assess how useful the average explanation is.
\end{itemize}

@@

@@

 \begin{figure}[t]
    \centering
    \includegraphics[width=0.75\linewidth]{images/Picture1.png}
    \caption{A high level building-block of our proposed hybrid fake voice identification model, where \(m\) and \(k\) represent the dimensionality of the given input feature.}
    \label{fig:overview}
\end{figure}
\section{Method}\label{sec:method}
Our proposed fake voice identification method consists of three major components including feature extractor, hybrid detection model, and the generation of explanations by modified LIME. After extracting features, we pass them individually through sub-models to overcome the multidimensionality problem. Then the concatenation of the outputs from the output layers of the sub-models are then passed through the \textit{terminus} models. For generating explanations, we make use of the feature values returned from the output layers of the sub-models, which has same representation of every feature.

@@

@@

\subsection{Hybrid Fake Voice Detection Model}

Fig.~\ref{fig:overview} depicts the overall architecture of our hybrid deep learning model for fake voice identification. We propose an architecture for our hybrid fake voice detection model that allows for a combined classification based on features of different dimensionality, or in other words, regardless of whether the individual features have the same shape. To achieve this, we hypothesize that we can make use of individual separate models for each feature, which we will further refer to as \textit{sub-models}. The main objective for having the individual sub-models is to have similar representation of the features so that we can use them for generating understandable explanations. The structure of sub-model can be different based the dimension of the extracted features. For example, the dimension of convolution layers and max pooling layer would be different based on the dimension of the features. Sub-models may also omit the pooling layers all-together, as depicted in the example in Fig.~\ref{fig:alt_model_arch} (a), in order to have a structure based purely on convolution. We hypothesize that the use of convolution within these sub-models will increase the localized pattern detection capability of our method.

After the processing within the sub-models, we then needed to further process and distill the results into a singular output. To achieve this, followed by the concatenations of the output of all sub-models, we propose a terminus model. With the goal of maximizing performance, we tried various structures for the terminus models including single layer perceptron, convolutional neural networks and multi-layer perceptron. The figure (Fig.~\ref{fig:overview}) portrays a multi-layer terminus containing either convolutional layers hidden perceptron layers. However, a single perceptron layers is also possible and would look like a single dense layer after the flattened concatenation, as exemplified in Fig.~\ref{fig:alt_model_arch} (b). We discuss in the experiments section about our experimental settings.

\begin{figure}[htbp]
	\centering
    \subfloat[]{\includegraphics[width=0.25\textwidth]{images/alt_sub_model.png}} 
	\hfil
    \subfloat[]{\includegraphics[width=0.5\textwidth]{images/simple_term.png}} 
	\caption{More examples for sub-model and terminus architectures, that deviate from the examples in Figure~\ref{fig:overview}. a) An example of a possible architecture of a sub-model without the inclusion of pooling layers. b) A terminus model with a simple perceptron architecture.}
    \label{fig:alt_model_arch}
\end{figure}
